---
title: "Visualizr Housing and Mortgage Data"
output:
  rmdformats::readthedown:
    highlight: pygments
    code_folding: hide
---

<style type="text/css">

p{ /* Normal  */   font-size: 14px;  line-height: 18px;}
body{ /* Normal  */   font-size: 14px;}
td {  /* Table  */   font-size: 12px;}
h1 { /* Header 1 */ font-size: 26px;
 color: #4294ce;}
h2 { /* Header 2 */ font-size: 22px;}
h3 { /* Header 3 */ font-size: 18px;}
code.r{ /* Code block */  font-size: 12px;}
pre { /* Code block */  font-size: 12px}

#table-of-contents h2 {background-color: #4294ce;}
#table-of-contents{background: #688FAD;}
#nav-top span.glyphicon{color: #4294ce;}
#postamble{background: #4294ce;border-top: ;}
</style>


```{r echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("tidyverse", "haven", "data.table", "scales", "ggtern", "tricolore", "cowplot",  prompt = FALSE)

setwd("~/R/Complete")
```

# Introduction

The future of the U.S. housing market depends on the evolution of demographic forces. There is a large group of young adults who will drive the U.S. housing market in years to come, but the U.S. also has the aging Baby Boomer generation.

What's going to happen as Boomers age? Typically as they age people move out of single family (SF) homes. But longer healthier lives might change the timing.

# Get data

PUMS - __P__ublic __U__se __M__icrodata __S__ample
In order to explore these trends, we're going to need to get some data. As we want custom tabulations, we cannot rely on the [ACS package](https://cran.r-project.org/web/packages/acs/index.html). Instead we'll use [IPUMS-USA](https://usa.ipums.org/usa/index.shtml) to download data from the [U.S. Census Bureau's Public Use Microdata Samples](https://www.census.gov/programs-surveys/acs/data/pums.html).

If you want to follow along you'll need to sign into IPUMS and download your own data. I'll share the details of the sample I'm using, but the file is too large to post here. I want to see how things may have shifted since 1980, so I'll collect samples from the 1980, 1990, and 2000 Decennial Census. I'll also grab the latest available microdata from the 2016 American Community Survey (and 2010 ACS as well). IPUMS has various options you can choose from when selecting your sample, but here are the variables I chose.

```{r loadData}
# load data ----
df <- read_dta("../Complete/data/usa_00001.dta")

#Made a mistake on IPUMS download.  Did not want 1930, did not need 2016

df <- filter(df, year > 1930, year != 2016)

# create variables
df <- mutate(df,
             own  = ifelse(ownershp==1,1,0),
             agec = cut(age,seq(0,100,5),right=FALSE),
             mtg  = case_when(
               mortgage == 0 ~ NA_real_,  # special NA for integers
               # https://stackoverflow.com/questions/44893933/avoiding-type-conflicts-with-dplyrcase-when
               mortgage == 1 ~ 0,
               TRUE ~ 1 )
             )
glimpse(df)
```

Now, we'll create two summary datasets. One will be based on households, the other persons. For all analysis we'll use individuals living outside of group quarters. We'll use households under the 1970s definition (GQ==1) so we can have consistency from 1980 forward. Though the additional counts you'd get from using the 1990s and 2000s household definitions are relatively minor.

```{r}
# household level dataframe ----

df.hha <-
  df %>% 
  filter(pernum==1 & gq %in% c(1)) %>%
  group_by(year, agec) %>%
  summarize(
            hh      = sum(hhwt), 
            renters = sum(ifelse(own==0,hhwt,0)),
            owners  = sum(ifelse(own==1,hhwt,0)),
            ho=weighted.mean(own,hhwt,na.rm=T),
            mtg= weighted.mean( 
              case_when(
                mortgage == 0 ~ NA_real_,
                mortgage == 1 ~ 0,
                TRUE ~ 1 ), 
              hhwt, na.rm=T),
            
            agemin = min(age,na.rm=T),
            agem = weighted.mean(age, hhwt,na.rm=T),
            agemax = max(age,na.rm=T),
            byrmin = min(birthyr, na.rm=T),
            byrmax = max(birthyr, nar.rm=T),
            mf= weighted.mean(ifelse(unitsstr %in% c(7,8,9,10),1,0),hhwt,na.rm=T),
            sf=weighted.mean(ifelse(unitsstr %in% c(3,4),1,0),hhwt,na.rm=T)
            ) %>%
  mutate( own.mtg = mtg*owners,
          own.clr = (1-mtg)*owners) %>%
  arrange(year, agec)

# person level dataframe ----
# won't use this one today, but could be useful if we wanted to do some person-level tabulations

df.pera <-
  df %>% 
  filter( gq %in% c(1)) %>%
  group_by(year, agec) %>%
  summarize(
            pop      = sum(perwt), 
            renters = sum(ifelse(own==0,perwt,0)),
            owners  = sum(ifelse(own==1,perwt,0)),
            ho=weighted.mean(own,perwt,na.rm=T),
            mtg= weighted.mean( 
              case_when(
                mortgage == 0 ~ NA_real_,
                mortgage == 1 ~ 0,
                TRUE ~ 1 ), 
              perwt, na.rm=T),
            agemin = min(age,na.rm=T),
            agem = weighted.mean(age, perwt,na.rm=T),
            agemax = max(age,na.rm=T),
            byrmin = min(birthyr, na.rm=T),
            byrmax = max(birthyr, nar.rm=T),
            mf= weighted.mean(ifelse(unitsstr %in% c(7,8,9,10),1,0),perwt,na.rm=T),
            sf=weighted.mean(ifelse(unitsstr %in% c(3,4),1,0),perwt,na.rm=T)
          ) %>%
  mutate( own.mtg = mtg*owners,
          own.clr = (1-mtg)*owners) %>%
  arrange(year, agec)
```

# Our Own Colors

For most of the plots below, we'll use a custom scale. We'll follow this post by Simon Jackson that describes how to create custom color palettes for ggplot2.

```{r}
 # Function for colors ----
  #####################################################################################
  ## Make Color Scale ----  ##
  #####################################################################################
  my_colors <- c(
    "green"      = rgb(103,180,75, maxColorValue = 256),
    "green2"      = rgb(147,198,44, maxColorValue = 256),
    "lightblue"  =  rgb(9, 177,240, maxColorValue = 256),
    "lightblue2" = rgb(173,216,230, maxColorValue = 256),
    'blue'       = "#00aedb",
    'red'        = "#d11141",
    'orange'     = "#f37735",
    'yellow'     = "#ffc425",
    'gold'       = "#FFD700",
    'light grey' = "#cccccc",
    'purple'     = "#551A8B",
    'dark grey'  = "#8c8c8c")
  
  
  my_cols <- function(...) {
    cols <- c(...)
    if (is.null(cols))
      return (my_colors)
    my_colors[cols]
  }
  
  my_palettes <- list(
    `main`  = my_cols("blue", "green", "yellow"),
    `cool`  = my_cols("blue", "green"),
    `hot`   = my_cols("yellow", "orange", "red"),
    `mixed` = my_cols("lightblue", "green", "yellow", "orange", "red"),
    `mixed2` = my_cols("lightblue2","lightblue", "green", "green2","yellow","gold", "orange", "red"),
    `mixed3` = my_cols("lightblue2","lightblue", "green", "yellow","gold", "orange", "red"),
    `mixed4` = my_cols("lightblue2","lightblue", "green", "green2","yellow","gold", "orange", "red","purple"),
    `mixed5` = my_cols("lightblue","green", "green2","yellow","gold", "orange", "red","purple","blue"),
    `mixed6` = my_cols("green", "gold", "orange", "red","purple","blue"),
    `grey`  = my_cols("light grey", "dark grey")
  )
  
  my_pal <- function(palette = "main", reverse = FALSE, ...) {
    pal <- my_palettes[[palette]]
    
    if (reverse) pal <- rev(pal)
    
    colorRampPalette(pal, ...)
  }
  
  scale_color_mycol <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
    pal <- my_pal(palette = palette, reverse = reverse)
    
    if (discrete) {
      discrete_scale("colour", paste0("my_", palette), palette = pal, ...)
    } else {
      scale_color_gradientn(colours = pal(256), ...)
    }
  }
  
  scale_fill_mycol <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
    pal <- my_pal(palette = palette, reverse = reverse)
    
    if (discrete) {
      discrete_scale("fill", paste0("my_", palette), palette = pal, ...)
    } else {
      scale_fill_gradientn(colours = pal(256), ...)
    }
  }
```

# Housing Trends

Let's compare trends in homeownership and households from 1980 to 2016 using our data.

```{r warning=FALSE}
mycaption <- "Decennial Census (1980,1990,2000) and 2010 and 2016 American Community Survey\nAuthor's tabulation of public use microdata via IPUMS-USA, University of Minnesota, www.ipums.org.\nRestricted to persons and households based on the 1970s household definition"

ggplot(summarize(df.hha, ho = weighted.mean(ho,hh)), aes(x=year,y=ho))+
  geom_path() + geom_point(size=3)+
  scale_x_continuous(breaks= unique(df.hha$year))+
  scale_y_continuous(labels=percent, breaks=seq(0,1,.01), limits=c(.63,.67))+
  theme_minimal(base_size=18)+
  theme(legend.position="none", plot.caption=element_text(hjust=0, size=8))+
  labs(x="Census Year", y="Homeownership Rate(%)", title="U.S. Homeownership Rate", caption=mycaption)
```

## Ownership by Age

We see that the homeownership rate in the United States is down a little in 2016. How does that compare by age group? Let's use our 5-year age buckets.

```{r}
ggplot(data=filter(df.hha,agemax<85,agemax>19), aes(x=year,y=ho, fill=agec))+geom_path() + geom_point()+
  scale_x_continuous(breaks= unique(df.hha$year)) + scale_y_continuous(labels=percent, breaks=seq(0,1,.1))+
  geom_area(alpha=0.25, position="identity") + theme_minimal(base_size=14) +
  scale_color_mycol("mixed6") + scale_fill_mycol("mixed6") + 
  theme(legend.position="none", axis.text=element_text(size=7),
        plot.caption=element_text(hjust=0, size=8), panel.spacing = unit(1, "lines")) +
  facet_wrap(~agec, ncol=5) + labs(x="Census Year", y="Homeownership Rate(%)", 
       title="U.S. Homeownership Rate by Householder Age", caption=mycaption)
```

Here we see an interesting pattern. The homeownership rate for the younger age groups have declined while they have generally risen for older cohorts. How about mortgage utilization? What's the percent of homeowner households that have a mortgage?

## Mortgage Utilization by Age

```{r}
ggplot(data=filter(df.hha,agemax<85,agemax>19), aes(x=year,y=mtg, fill=agec, color=agec))+
  geom_path(color="black") + geom_point() +
  scale_x_continuous(breaks= unique(df.hha$year)) + scale_y_continuous(labels=percent, breaks=seq(0,1,.1))+
  geom_area(alpha=0.25, position="identity", color="NA") + theme_minimal(base_size=14) +
  scale_color_mycol("mixed6") + scale_fill_mycol("mixed6") +
  theme(legend.position="none", axis.text=element_text(size=7), plot.caption=element_text(hjust=0, size=8),
        panel.spacing = unit(1, "lines")) + facet_wrap(~agec, ncol=5) +
  labs(x="Census Year", y="Percent of homeowners with mortgage(%)", 
       title="U.S. Mortgage utilization by householder age",
       subtitle="Percent of homeowner households with a mortgage", caption=mycaption)
```

Here we see that over age, the share of households with a mortgage tends to decline, though the share of older households with a mortgage has increased. Let's combine rentership with two types of owners, those with a mortgage and those who own free and clear in 2016. We'll need to do a tiny bit of data manipulation.

## Tenure & Mortgage Status by Age

```{r}
# roll up by year with group_by and summarize
df.bar <- df.hha %>% filter( agem< 85) %>% group_by(year,agec) %>%
  summarize(renter=sum(renters, na.rm=T), own.with.mtg = sum(hh*ho*mtg, na.rm=T), 
            own.clear = sum(hh*ho*(1-mtg), na.rm=T)) %>% gather(tenure,hh,-year,-agec)

ggplot(data=filter(df.bar, year == 2017), aes(x = agec, y = hh, fill = tenure)) + 
  geom_col(position = "stack", alpha = 0.82) + scale_color_mycol("mixed6") + scale_fill_mycol("mixed6")+
  theme_minimal(base_size = 14) + scale_y_continuous(labels = comma,expand=c(0,0))+
  theme(legend.position = "top", axis.text = element_text(size = 7), 
        plot.caption = element_text(hjust = 0, size = 8), panel.spacing = unit(1, "lines")) +
  labs(x = "Householder Age", y = "Number of households", title = "U.S. households in 2016",
       subtitle = "Tenure and mortgage status of households by age", caption = mycaption)
```

If we were feeling mischievous we could use a polar coordinate/radial chart. First with the bars (now wedges) summing up to 1 (share of total):

```{r out.height=10, out.width=8}
ggplot(data=df.bar, aes(x=agec, y=hh,fill=tenure)) + geom_col(position = "fill", alpha = 0.82)+
  scale_color_mycol("mixed6") + scale_fill_mycol("mixed6") + theme_minimal(base_size = 14) +
  coord_polar() + facet_wrap(~year) +
  theme(legend.position = "top", axis.text = element_text(size = 7), plot.caption = element_text(hjust = 0, size = 8),
        panel.spacing = unit(1, "lines")) +
  labs(x = "Householder Age", y = "Share of households", 
       title = "U.S. households: renters and owners with and without a mortgage",
       subtitle = "Tenure and mortgage status of households by age", caption = mycaption)
```

Or in absolute counts, wrapping the stacked bar up above around a circle.

```{r out.height=10, out.width=8}
ggplot(data=df.bar, aes(x=agec, y=hh,fill=tenure)) + geom_col(position="stack", alpha=0.82)+
  scale_color_mycol("mixed6") + scale_fill_mycol("mixed6") + theme_minimal(base_size=14)+
  coord_polar() + facet_wrap(~year) +
  theme(legend.position="top", axis.text=element_text(size=7), plot.caption=element_text(hjust=0, size=8),
        panel.spacing = unit(1, "lines"))+
  labs(x="Householder Age", y="Number of households", 
       title="U.S. households: renters and owners with and without a mortgage",
       subtitle="Tenure and mortgage status of households by age", caption=mycaption)
```

Trace out the share of households living in apartments over time:

```{r message=FALSE}
ggplot(data=filter(df.hha, agemax>19 , agemax<85 ) %>% arrange(year,agemax), aes(x=agec,y=mf,color=factor(year),linetype=factor(year), group=year))+  geom_path(size=1.05)+
  scale_color_mycol("mixed6") + scale_fill_mycol("mixed6") + theme_minimal(base_size=14)+
  geom_text(data=. %>% filter(agemax==84), aes(label=year), hjust=0,nudge_x=0.1,size=4)+
  theme(legend.position="none", axis.text=element_text(size=7), plot.caption=element_text(hjust=0, size=8),
        panel.spacing = unit(1, "lines")) + scale_y_continuous(labels=percent) +
  labs(title="Percent of U.S. households living in multifamily housing \n(structures with 5+ housing units)", x="Householder Age",y="Percent of households", caption=mycaption)
```
This chart reveals that the U shaped pattern we saw above is consistent over time. However, the curve has shifted. Interesintly it has shifted up for ages less than 60, but down for those over 60. Could it shift down further in future? It might as Boomers age and live longer healthier lives than earlier generations.
We can unpack the plot and use facets to examine how age groups have evolved over time.

```{r}
ggplot(data=filter(df.hha,agemax<85,agemax>19), aes(x=year,y=mf, fill=agec, color=agec))+
  geom_path(color="black") + geom_point() + scale_x_continuous(breaks= unique(df.hha$year))+
  scale_y_continuous(labels=percent, breaks=seq(0,1,.1)) + 
  geom_area(alpha=0.25, position="identity", color="NA") +
  theme_minimal(base_size=14) + scale_color_mycol("mixed6") + scale_fill_mycol("mixed6")+
  theme(legend.position="none", axis.text=element_text(size=7),
        plot.caption=element_text(hjust=0, size=8), panel.spacing = unit(1, "lines"))+
  facet_wrap(~agec, ncol=5) +
  labs(x="Census Year", y="Percent of Households (%)", 
       title="Percent of U.S. Households living in multifamily (5+ unit)\nstructures by householder age",
       # subtitle="Percent of homeowner households with a mortgage",
       caption=mycaption)
```

Tricolore
This is worth exploring in more detail in a future post, but I just learned about the very interesting Tricolore package. This allows you to make charts with a flexible color scale for the visualization of three-part/ternary compositions. It's like taking bivariate color scales to a whole new dimension.
We happen to have a nice example, with the three types of households we reviewed above (renters, owners with a mortgage, and owners who own free and clear). If we install the package via devtools::install_github('jschoeley/tricolore, we can run the following and get:

```{r out.height=14, warning=FALSE}

# Might need to run this to avoid breaking ggtern with cowplot
# see https://github.com/wilkelab/cowplot/issues/88 
theme_set(theme_gray())

# create complex plot

hha.test <- filter(df.hha,agemin< 85, agemin>19) %>% mutate(r=(1-ho), m=(1-ho)*mtg, f=(1-ho)*(1-mtg))

#df.tri <- Tricolore(hha.test, "r","m","f")

df.tri <- Tricolore(hha.test, "renters","own.mtg","own.clr")
#hha.test$srgb <- df.tri$hexsrgb  # merge colors back on data
hha.test$srgb <- df.tri$rgb  # merge colors back on data

g.tri <- df.tri$legend + theme(plot.background = element_rect(fill = NA, color = NA))

g.tile <- ggplot(data=hha.test, aes(x=agec, y=factor(year), fill=srgb)) +
  geom_tile(color="white") + labs(x="Householder Age", y="Year") +
  scale_fill_identity() + coord_flip()

# put triangle legend with tile plot
g3 <- plot_grid(g.tri,g.tile,rel_widths=c(1,2))

df.bar3 <- left_join(df.bar, select(hha.test, year,agec,srgb), by=c("year","agec"))

mycaption <- "Source: Decennial Census (1980,1990,2000) and 2010 and 2016 American Community Survey\nAuthor's tabulation of public use microdata via IPUMS-USA, University of Minnesota, www.ipums.org.\nRestricted to persons and households based on the 1970s definition"

g.bar <- 
  ggplot(data=filter(df.bar3,!is.na(srgb)), aes(x=agec, y=hh/1000,fill=srgb, group=tenure)) + 
  geom_col(position="stack", alpha=0.82) + scale_fill_identity()+
  theme_minimal(base_size=14) + facet_grid(year~tenure)+
  scale_y_continuous(labels=comma) + theme(legend.position="top",
        axis.text=element_text(size=8), plot.caption=element_text(hjust=0, size=8),
        panel.spacing = unit(1, "lines")) +
  labs(x="Householder Age", y="Number of households (1000s)", 
       title="U.S. households that rent, own with a mortgage and own free and clear",
       subtitle="Tenure and mortgage status of households by age",
       caption=paste0(mycaption,"\nTricolore color based on https://github.com/jschoeley/tricolore"))

# create combined plot
cowplot::plot_grid(g3,g.bar, ncol=1, rel_heights=c(1,2))
```


# Reference

All information provided by Len Kiefer from RStudio::conf 2019